Nonlinear mixed effects models for time series forecasting of smart meter demand

### 2020-10-11 paper.Rmd
Proofread. Fix typos.

The relationship between current temperature and electricity demand is shown in Figure \@ref(fig:temp-wh-scatter) for two buildings at midday and midnight. There is a clear difference in this relationship between the business and non-business periods. Furthermore, this relationship varies between each building. The different relationships for each building necessitates the use of a nonlinear demand response function that can adequately capture the unique response characteristics of each building. To do this, we model our data with mixed effects models and capture the nonlinear building specific demand response using subject-specific curves. Note that when splines are fitted independently for each building they can be very sensitive to outliers or data at domain extremes. This also motivates our use of mixed effects models where data from multiple buildings can be used to fit these relationships giving models that are more robust to outliers or sparse data at covariate extremes.

@@

We use natural splines with three degrees of freedom as our smooth functions. Knot positions $\kappa_k$ are calculated based on quantiles of the data. Natural splines are chosen over other types as they enforce the constraint of linearity beyond the boundary points, which seems a fair assumption when considering the behaviour of electricity demand consumption in relation to extreme temperatures (see Figure \@ref(fig:plot-linear-cubic-splines) for an illustration). We wish to create a parsimonious model and assuming anything beyond a linear relationship in the extremes seems contrary to that aim. Failing to enforce the linearity constraint may result in unusual relationships being predicted if extrapolating beyond the training data.

@@

Additional predictor variables are introduced in this model, such as lagged temperature variables; maximum, minimum and average temperatures; and lagged demand. When doing one day ahead forecasts we do not have have demand observations within the last 24 hours to use as lags and so we restrict our lagged demand variables to 1 day, 2 days and 3 days. The temperature variables can have 1-24 hour lags as temperature forecasts can be used to supplement the already observed data. Note that in this paper we restrict ourselves to using actual temperature data and not forecast temperature data to ensure our results are dependent on model formulations and not on any errors in temperature forecasts. In practice, forecast temperature data can be used instead. A description of each predictor is presented in Table \@ref(tab:weather-vars). We denote this set of predictor variables as $\mathcal{P}_t$, which contains the selected variables for the 15-minute period of day and month at time $t$. The exact combination of variables is chosen via our feature selection methodology described in Section \@ref(feature-selection). This model is used when selecting features as it is much faster to train than a mixed effects model.

>>>>

The relationship between current temperature and electricity demand is shown in Figure \@ref(fig:temp-wh-scatter) for two buildings at midday and midnight. There is a clear difference in this relationship between the business and non-business periods. Furthermore, this relationship varies between each building. The different relationships for each building necessitates the use of a nonlinear demand response function that can adequately capture the unique response characteristics of each building. To accomplish this, we model our data with mixed effects models and capture the nonlinear building specific demand response using subject-specific curves. Note that when splines are fitted independently for each building they can be very sensitive to outliers or data at domain extremes. This also motivates our use of mixed effects models where data from multiple buildings can be used to fit these relationships giving models that are more robust to outliers or sparse data at covariate extremes.

@@

We use natural splines with three degrees of freedom as our smooth functions. Natural splines are chosen over other types as they enforce the constraint of linearity beyond the boundary points, which seems a fair assumption when considering the behaviour of electricity demand consumption in relation to extreme temperatures (see Figure \@ref(fig:plot-linear-cubic-splines) for an illustration). We wish to create a parsimonious model and assuming anything beyond a linear relationship in the extremes seems contrary to that aim. Failing to enforce the linearity constraint may result in unusual relationships being predicted if extrapolating beyond the training data.

@@

Additional predictor variables are introduced in this model, such as lagged temperature variables; maximum, minimum and average temperatures; and lagged demand. When doing one day ahead forecasts we do not have demand observations within the last 24 hours to use as lags and so we restrict our lagged demand variables to 1 day, 2 days and 3 days. The temperature variables can have 1-24 hour lags as temperature forecasts can be used to supplement the already observed data. Note that in this paper we restrict ourselves to using actual temperature data and not forecast temperature data to ensure our results are dependent on model formulations and not on any errors in temperature forecasts. In practice, forecast temperature data can be used instead. A description of each predictor is presented in Table \@ref(tab:weather-vars). We denote this set of predictor variables as $\mathcal{P}_t$, which contains the selected variables for the 15-minute period of day and month at time $t$. The exact combination of variables is chosen via our feature selection methodology described in Section \@ref(feature-selection). This model is used when selecting features as it is much faster to train than a mixed effects model.
